---
title: "Global Baseline Predictors and RMSE"
author: "Vikas Sinha"
date: "June 11 2019"
output:
  html_document:
    code_folding: show
    df_print: paged
    number_sections: yes
    toc: yes
    toc_depth: 6
    toc_float: yes
  pdf_document: default
  word_document:
    toc: yes
    toc_depth: '6'
---


# Introduction

The following is an implementation of the Baseline Predictor for a Recommender System that recommends movies. This system is meant to demonstrate the usage of baseline predictors that compute biases for each user and item. The data set used is from *grouplens.org*.


```{r load, eval=T, echo=T, warning=F, message=F}
library(caret)
library(data.table)
library(dplyr)
library(earth)
library(ggplot2)
library(gridExtra)
library(ipred)
library(knitr)
library(lubridate)
library(mice)
library(ModelMetrics)
library(party)
library(psych)
library(rpart)
library(tidyr)
library(utils)
library(recommenderlab)
library(reshape2)
library(rrecsys)
```


# Dataset

The data used are extracted from the file *ml-latest-small.zip* downloaded from (http://grouplens.org/datasets/movielens/latest). GroupLens is a research lab at the University of Minnesota that specializes in recommendation systems and related research areas.

We start by examining summaries and the first few records of the dataframes corresponding to the movies and their ratings.

For the purpose of this assignment we reduce the ratings dataset to about 1000 ratings.


```{r Load, eval=T, echo=T, warning=F, message=F}
moviefile = "ml-latest-small/movies.csv"
ratingsfile = "ml-latest-small/ratings.csv"
movies = read.csv(moviefile)
ratings = read.csv(ratingsfile)
head(movies)
summary(movies)


# Reduce the size of the data set for this assignment.

set.seed(123)
reduced = sample.int(nrow(ratings), 1000, replace = F)
ratings = ratings[reduced,]
head(ratings)
summary(ratings)
```


Since the ratings file contains one entry per rating, we can calculate the number of unique users as follows.

```{r uniq1, eval=T, echo=T, warning=F, message=F}
cat("The number of users in the data set is:", length(unique(ratings$userId)))
```


# User-Item Matrix


A user-item matrix is created, where each row represents a user and each column a movie (item). The element for a given row and column represents the rating given by the user to the movie.

The *dcast()* function from the *data.table* library is used to convert from long to wide format.


```{r da1, eval=T, echo=T, warning=F, message=F}
ratings2 = dplyr::select(ratings, -timestamp)
ui = data.table::dcast(ratings2, userId ~ movieId, value.var = "rating")

ui2 = as.matrix(dplyr::select(ui, -userId))
kable(data.frame(Users=nrow(ui2), Movies=ncol(ui2)),
      caption="Total Size of User-Item matrix (Sparse)")
```


This table shows the frequency of NAs in the User-Item matrix.
We can see that the U-I matrix is very sparse, since the count of NA values is much greater than the count of valid values in that matrix.

  
```{r da2, eval=T, echo=T, warning=F, message=F}
table(is.na(ui2))
```



## Split into test/train subsets.


Split the data set into test and train subsets. To do this, we perform the following procedure:


- Create a Test matrix and set all of its elements to NA.
- Select a row at random from the *ratings* dataframe. Obtain the userId and movieId.
- Convert userId, movieId to strings (using as.character())
- Search for the element in the User-Item Matrix that matches the given userId, movieId, and
    - set its value to NA for the train matrix
    - set its value to the actual value for the test matrix


```{r tt1, eval=T, echo=T, warning=F, message=F}
set.seed(124)
rows = sample.int(nrow(ratings), size = 0.3*nrow(ratings), replace = F)

r.train = ratings2
#r.test = ratings2
r.test = setNames(data.frame(matrix(ncol = 3, nrow = 0)), colnames(ratings2))


jj = 1
for (ii in rows) {
  
  #r.test[jj,]$userId = ratings2[ii,]$userId
  #r.test[jj,]$movieId = ratings2[ii,]$movieId
  #r.test[jj,]$rating = ratings2[ii,]$rating
  #jj = jj + 1

  r.test[jj,] = ratings2[ii,]
  jj = jj + 1

  r.train[ii,]$rating = NA
}

```


# Raw ratings.

Raw (mean) rating for each user and item combination.
In a raw average, every entry gets the same prediction, which is the mean over the data set.

```{r m1, eval=T, echo=T, warning=F, message=F}
raw.rating = sum(ratings2$rating) / nrow(ratings2)

cat("The raw (global) rating for the data set is:", raw.rating)
```


Calculate RMSE for the raw average.

```{r rmse1, eval=T, echo=T, warning=F, message=F}
r.test$raw.rating = raw.rating

cat("The raw average RMSE is:", rmse(r.test$raw.rating, r.test$rating))
```



# Biases for users and items.

Calculate bias for each user and for each item.


```{r eval1, eval=T, echo=T, warning=F, message=F}
UserBias <- function(uid, df) {
  ss = df[which(df$userId == uid),]
  ub = sum(ss$rating)/nrow(ss)
  return(ub)
}

ItemBias <- function(mid, df) {
  ss = df[which(df$movieId == mid),]
  ib = sum(ss$rating)/nrow(ss)
  return(ib)
}


```


# RMSE for Baseline Predictor

```{r rmse2, eval=T, echo=T, warning=F, message=F}
r.test$baseline = 0.0

for (ii in 1:nrow(r.test)) {
  r.test[ii,]$baseline = r.test[ii,]$raw.rating +
                           UserBias(r.test[ii,]$userId, ratings2) - raw.rating +
                           ItemBias(r.test[ii,]$movieId, ratings2) - raw.rating
}

cat("The RMSE for the baseline predictor is:", rmse(r.test$baseline, r.test$rating))
```


# References

1. MovieLens Latest Datasets. https://grouplens.org/datasets/movielens/latest/
2. F. Maxwell Harper and Joseph A. Konstan. 2015. The MovieLens Datasets: History and Context. ACM Transactions on Interactive Intelligent Systems (TiiS) 5, 4: 19:1–19:19. https://doi.org/10.1145/2827872
