---
title: "Recommender System Using Baseline Predictor"
author: "Vikas Sinha"
date: "June 10 2019"
output:
  html_document:
    code_folding: show
    df_print: paged
    number_sections: yes
    toc: yes
    toc_depth: 6
    toc_float: yes
  pdf_document: default
  word_document:
    toc: yes
    toc_depth: '6'
---


# Introduction

The following is an implementation of a Recommender System that recommends movies. This system is meant to demonstrate the usage of baseline predictors that compute biases for each user and item. The data set used is from *grouplens.org*.


```{r load, eval=T, echo=T, warning=F, message=F}
library(caret)
library(data.table)
library(dplyr)
library(earth)
library(ggplot2)
library(gridExtra)
library(ipred)
library(knitr)
library(lubridate)
library(mice)
library(party)
library(psych)
library(rpart)
library(tidyr)
library(utils)
library(recommenderlab)
library(reshape2)
library(rrecsys)
```


# Dataset

GroupLens is a research lab at the University of Minnesota that specializes in recommendation systems. The data used are in the file *ml-latest-small.zip* downloaded from the URL (http://grouplens.org/datasets/movielens/latest).

We next examine the summaries of the dataframes corresponding to the movies and their ratings.

```{r Load, eval=T, echo=T, warning=F, message=F}
moviefile = "ml-latest-small/movies.csv"
ratingsfile = "ml-latest-small/ratings.csv"

movies = read.csv(moviefile)
ratings = read.csv(ratingsfile)

head(movies)
summary(movies)

head(ratings)
summary(ratings)
```


Since the ratings file contains one entry per rating, we can find the number of unique users as follows.

```{r uniq1, eval=T, echo=T, warning=F, message=F}
cat("The number of unique users in the data set is:", length(unique(ratings$userId)))
```


# User-Item Matrix

A user-item matrix is created, with each row representing a user and each column representing a movie (item). The element for a given row and column represents the rating given by the user to the movie. The *dcast()* function is used to convert from long to wide format.

```{r da1, eval=T, echo=T, warning=F, message=F}
ratings2 = dplyr::select(ratings, -timestamp)
ui = data.table::dcast(ratings2, userId ~ movieId, value.var = "rating")
ui2 = as.matrix(dplyr::select(ui, -userId))
```



## Test/Train data sets

Split the data set into test and train portions.

```{r tt1, eval=T, echo=T, warning=F, message=F}
```


# Raw (mean) rating for each user and item combination. Calculate RMSE for the raw average.

what is baseline predictor?
raw average: every entry gets the same prediction, mean over the data set.

```{r m1, eval=T, echo=T, warning=F, message=F}

```


# Calculate bias for each user and for each item.

raw average and rmse calculation

```{r eval1, eval=T, echo=T, warning=F, message=F}

```


# References

1. MovieLens Latest Datasets. https://grouplens.org/datasets/movielens/latest/
2. F. Maxwell Harper and Joseph A. Konstan. 2015. The MovieLens Datasets: History and Context. ACM Transactions on Interactive Intelligent Systems (TiiS) 5, 4: 19:1–19:19. https://doi.org/10.1145/2827872
