---
title: "Recommender System For Hotel Recommendations"
author: "Vikas Sinha"
date: "July 14 2019"
output:
  pdf_document: default
  html_document:
    code_folding: hide
    df_print: paged
    number_sections: yes
    toc: yes
    toc_depth: 6
    toc_float: yes
  word_document:
    toc: yes
    toc_depth: '6'
---


# Introduction

This is an implementation of a system for recommending hotel reservations.
The problem description and data are based on a Kaggle competition defined by Expedia.

The problem description is available on Kaggle [1] and the data set may be downloaded from:
https://www.kaggle.com/c/expedia-hotel-recommendations/data


# Recommendation Pipeline

The pipeline for the recommendation system implemented here has two main components:

- In the first step, a classifier is built to predict the cluster ID of the hotel
cluster that is preferred by a user, based on the input features in the train set. The
classifier used here is a Gradient Boosting Machine.

- Given the features in the test data set, a prediction is made using the
  GBM classifier of Step 1. The hotel cluster predicted is then input
  to an item similarity algorithm in order to predict a set of similar clusters that
  the user would be interested in. This set of similar items is the output
  of the recommender system.

For cluster prediction in Step 1 we use a Gradient Boosting Machine from the library
*gbm*.
The response variable here is the *hotel_cluster* field in the train data set.

For item similarity computation we use the *lsa* library in R to calculate
cosine similarity based on the array of latent features provided in the data set.


# Libraries

The following R libraries are used.

```{r load, eval=T, echo=T, warning=F, message=F}
library(class)
library(cluster)
library(dplyr)
library(gbm)
library(ggplot2)
library(knitr)
library(lsa)
library(reshape2)
library(tibble)

(.packages())
```



# Data Exploration


We load the CSV files train.csv and destinations.csv.
To enable processing on a single node, we limit the number of
rows to 1000000. Since we are only interested in those observations
that led to a sale (reservation) and not in those that were mere clicks,
we filter out observations that did not result in a booking.


```{r load1, eval=T, echo=T, warning=F, message=F}
tdf = read.csv("expedia/train.csv", nrow=1000000)
tdf = na.omit(tdf) %>% dplyr::filter(is_booking == 1)

latent_features_df = read.csv("expedia/destinations.csv")
tdf$site_name = as.factor(tdf$site_name)

set.seed(202)
train_index = sample.int(nrow(tdf), size = floor(0.75*nrow(tdf)), replace=F)
train = tdf[train_index,]
test = tdf[-train_index,]

head(tdf)
```

These are the dimensions of the dataframe containing latent features
of destinations.

```{r lfd, eval=T, echo=T, warning=F, message=F}
dim(latent_features_df)
```

We now explore some of the information included in the training data set.
These are the features present in the data set with their text descriptions provided by
Expedia.


```{r exp1, eval=T, echo=T, warning=F, message=F}
description = c(
    "Timestamp",
    "ID of the Expedia point of sale (i.e. Expedia.com, Expedia.co.uk, ...)",
    "ID of continent associated with site_name",
    "The ID of the country the customer is located",
    "The ID of the region the customer is located",
    "The ID of the city the customer is located",
    "Physical distance between a hotel and acustomer at the time of search",
    "ID of user",
    "1 when a user connected from a mobile device",
    "1 if the click/booking was generated as a part of a package",
    "ID of a marketing channel",
    "Checkin date   string",
    "Checkout date  string",
    "The number of adults specified in the hotel room",
    "The number of (extra occupancy) children specified",
    "The number of hotel rooms specified in the search",
    "ID of the destination where the hotel search was performed",
    "Type of destination",
    "1 if a booking, 0 if a click",
    "Numer of similar events in the context of the same user session",
    "Hotel continent",
    "Hotel country",
    "Hotel market",
    "ID of a hotel cluster")

desc = data.frame(Feature=names(train), Description=description)
kable(desc)
```

We can see that the following features are present in the train data but not in the
test data:

- "is_booking"
- "cnt"
- "hotel_cluster"

Also, the test data has the field *id* which is not present in the train data
and seems to be just an index field that can be removed.

The file *destinations.csv* contains latent features.

Data fields *destinations.csv*
d1-d149     latent description of search regions

# Predictive Model for Hotel Selection

We now build a Gradient Boosting Machine model to predict hotel clusters based on a selection of input features
provided in the data files. For this, only those entries that have the field *is_booking* set to 1 are used, in
order to allow the trained model to predict destinations when the user would be likely to make a purchase.


```{r sp1, eval=T, echo=T, warning=F, message=F}
set.seed(234)
gbm.train = train
gbm.train = dplyr::select(gbm.train, hotel_cluster,
                          site_name, user_location_country,
                          user_location_region, user_location_city,
                          orig_destination_distance,
                          hotel_country,
                          hotel_market,
                          srch_destination_id,
                          srch_children_cnt,
                          channel,
                          srch_rm_cnt)

m.gbm = gbm(hotel_cluster ~ .,
            distribution = "gaussian",
            data = gbm.train)
print(m.gbm)

gbm.test = dplyr::select(test, -hotel_cluster)
ypred = predict(m.gbm, n.trees=m.gbm$n.trees, test)
```

This is the RMSE of the prediction of hotel clusters, as computed
against the test portion of the data set.

```{r err1, eval=T, echo=T, warning=F, message=F}
caret::RMSE(ypred, as.integer(test$hotel_cluster))
```


In this section we compute similarity of hotel clusters by
computing cosine similarity on the provided matrix of latent features.
The list of similarity values is sorted and the top N values,
corresponding to the most similar items, are returned. For this
the *lsa* package's *cosine* function is used.


```{r sim1, eval=T, echo=T, warning=F, message=F}
# Arguments:
# id: id of the hotel cluster for which similar set is to be computed
# lfa_df: dataframe of latent features.

get_similar_items <- function(id, lfa_df) {
    rows = nrow(lfa_df)
    s = as.numeric(lfa_df[id,])
    siminfo = numeric(rows)
    for (ii in 1:rows) {
        siminfo[ii] = cosine(s, as.numeric(lfa_df[ii,]))
    }

    # Return the top 10 similar items
    s = sort(siminfo, decreasing=T)
    s = head(s, 10)
    return(s)
}

```


## Computing Similar Items

Here we show an example of how the similarity function based
on latent features is used to return similar items. The function
*get_similar_items()* implemented above is invoked to return
a list of most similar items.


```{r sim2, eval=T, echo=T, warning=F, message=F}
id = as.integer(ypred[1])
similar = get_similar_items(id, latent_features_df)
```


# Conclusions

This project is an implementation of a recommender system for
hotel recommendations based on click data related to users' online behavior.
The response variable is the identifier of the hotel cluster selected by
the user.

The recommender strategy implemented here is to first predict the response variable
based using a Gradient Boosting Machine trained on the subset of provided
data for which users actually made purchases. The GBM model's output is then
used, in conjunction with provided data on latent features of hotel clusters,
to compute a set of similar clusters that the user would be interested in.

We used a portion of the training data provided to evaluate the algorithm
and reported the RMSE achieved.


# References

1. Expedia Hotel Recommendations. https://www.kaggle.com/c/expedia-hotel-recommendations/overview
2. https://towardsdatascience.com/a-machine-learning-approach-building-a-hotel-recommendation-engine-6812bfd53f50
3. Gradient Boosting Machine. http://uc-r.github.io/gbm_regression
